---
title: "Batch correct Wagner dataset, plot heatmaps"
output:
  html_document:
    keep_md: true
    smart: false
    theme: united
date: 'Compiled: `r format(Sys.Date(), "%B %d, %Y")`'
---
***

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache.lazy = FALSE,
  tidy = TRUE
)
```

## Libraries
```{r}
suppressMessages(library(plotly))
suppressMessages(library(Seurat))
suppressMessages(library(dplyr))
suppressMessages(library(Matrix))
suppressMessages(library(gprofiler2))
suppressMessages(library(gplots))
suppressMessages(library(future))
suppressMessages(library(SeuratWrappers))
suppressMessages(library(SeuratDisk))
suppressMessages(library(gridExtra))
suppressMessages(library(ggplot2))
suppressMessages(library(ggrepel))

outputDir = getwd()

# A list of cell cycle markers, from Tirosh et al, 2015, is loaded with Seurat.  We can
# segregate this list into markers of G2/M phase and markers of S phase
s.genes <- cc.genes$s.genes
g2m.genes <- cc.genes$g2m.genes
```

# Wagner dataset
* Concerns ovary tissue from c-sections and gender reassignment patients (GRP)
* DOI: 10.1038/s41467-020-14936-3
* We only use non sorted 10X data

## Load the wagner data, Create Seurat objects
```{r}
c_sec_data <- Read10X_h5(filename = "/path/to/filtered_gene_bc_matrices_h5.h5")
grp_data <- Read10X_h5(filename = "/path/to/filtered_gene_bc_matrices_h5.h5")

c_sec <-  CreateSeuratObject(counts = c_sec_data, 
                             min.cells = 3, 
                             min.features = 100,
                             project = "c_sec")
grp <-  CreateSeuratObject(counts = grp_data, 
                           min.cells = 3, 
                           min.features = 100,
                           project = "grp")
```

## Merge the objects
```{r}
wagner <- merge(x = c_sec, 
                y = grp,
                add.cell.ids = c("c-sec", "gpr"))
```

# Seurat workflow
## QC
* In the paper they discard cells expressing over 7000 genes. Probably too lenient, we discard above 5000
* In the paper they use a 25% MT threshold. Probably too lenient, we choose 10%
```{r}
wagner[["percent.mt"]] <- PercentageFeatureSet(wagner, pattern = "^MT-")
VlnPlot(wagner, 
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), 
        ncol = 3)
wagner <- subset(wagner, 
                    subset = nFeature_RNA > 200 & nFeature_RNA < 5000 & percent.mt < 10)
VlnPlot(wagner, 
        features = c("nFeature_RNA", "nCount_RNA", "percent.mt"), 
        ncol = 3)
table(wagner$orig.ident)
```

## Normalize the data
* Notice we use a scale factor of 20000, to be on the upper side of the cell read count
```{r}
wagner <- NormalizeData(wagner, 
                        normalization.method = "LogNormalize", 
                        scale.factor = 20000)
```

## Dissociation genes on normalized data
* Check the expression of dissociation genes compared to total expression (for details see 10.1038/nmeth.4437)
```{r, message=FALSE}
genesChrom <- c("Actg1__chr11","Ankrd1__chr19","Arid5a__chr1","Atf3__chr1","Atf4__chr15","Bag3__chr7","Bhlhe40__chr6",
"Brd2__chr17","Btg1__chr10","Btg2__chr1","Ccnl1__chr3","Ccrn4l__chr3","Cebpb__chr2","Cebpd__chr16",
"Cebpg__chr7","Csrnp1__chr9","Cxcl1__chr5","Cyr61__chr3","Dcn__chr10","Ddx3x__chrX","Ddx5__chr11",
"Des__chr1","Dnaja1__chr4","Dnajb1__chr8","Dnajb4__chr3","Dusp1__chr17","Dusp8__chr7",
"Egr1__chr18","Egr2__chr10","Eif1__chr11","Eif5__chr12","Erf__chr7","Errfi1__chr4","Fam132b__chr1",
"Fos__chr12","Fosb__chr7","Fosl2__chr5","Gadd45a__chr6","Gcc1__chr6","Gem__chr4","H3f3b__chr11",
"Hipk3__chr2","Hsp90aa1__chr12","Hsp90ab1__chr17","Hspa1a__chr17","Hspa1b__chr17","Hspa5__chr2",
"Hspa8__chr9","Hspb1__chr5","Hsph1__chr5","Id3__chr4","Idi1__chr13","Ier2__chr8","Ier3__chr17",
"Ifrd1__chr12","Il6__chr5","Irf1__chr11","Irf8__chr8","Itpkc__chr7","Jun__chr4","Junb__chr8",
"Jund__chr8","Klf2__chr8","Klf4__chr4","Klf6__chr13","Klf9__chr19","Litaf__chr16","Lmna__chr3",
"Maff__chr15","Mafk__chr5","Mcl1__chr3","Midn__chr10","Mir22hg__chr11","Mt1__chr8","Mt2__chr8",
"Myadm__chr7","Myc__chr15","Myd88__chr9","Nckap5l__chr15","Ncoa7__chr10","Nfkbia__chr12","Nfkbiz__chr16",
"Nop58__chr1","Nppc__chr1","Nr4a1__chr15","Odc1__chr12","Osgin1__chr8","Oxnad1__chr14","Pcf11__chr7",
"Pde4b__chr4","Per1__chr11","Phlda1__chr10","Pnp__chr14","Pnrc1__chr4","Ppp1cc__chr5","Ppp1r15a__chr7",
"Pxdc1__chr13","Rap1b__chr10","Rassf1__chr9","Rhob__chr12","Rhoh__chr5","Ripk1__chr13","Sat1__chrX",
"Sbno2__chr10","Sdc4__chr2","Serpine1__chr5","Skil__chr3","Slc10a6__chr5","Slc38a2__chr15",
"Slc41a1__chr1","Socs3__chr11","Sqstm1__chr11","Srf__chr17","Srsf5__chr12","Srsf7__chr17",
"Stat3__chr11","Tagln2__chr1","Tiparp__chr3","Tnfaip3__chr10","Tnfaip6__chr2","Tpm3__chr3",
"Tppp3__chr8","Tra2a__chr6","Tra2b__chr16","Trib1__chr15","Tubb4b__chr2","Tubb6__chr18",
"Ubc__chr5","Usp2__chr9","Wac__chr18","Zc3h12a__chr4","Zfand5__chr19","Zfp36__chr7","Zfp36l1__chr12",
"Zfp36l2__chr17","Zyx__chr6","Gadd45g__chr13","Hspe1__chr1","Ier5__chr1","Kcne4__chr1")

genes <- sapply(genesChrom, function(x){
  toupper( strsplit(x, "__")[[1]][1])
})

Data <- as.data.frame(as.matrix(GetAssayData(object = wagner)))
cat("All genes:\n")
print(unname(genes))
write.table(genes, paste0(outputDir, "/mouseDissocGenes.tsv"), sep ="\t", quote=FALSE, row.names = FALSE)

## Remove mouse only genes and put the corresponding human
genes <- genes[!genes %in% c("CCRN4L", "MT1", "MT2")]
genes <- c(genes, "NOCT", "MT1A", "MT2A")
cat("Genes from mouse we miss in human:\n")
unname(genes[!genes %in% row.names(Data)])

## Calculate the percentage of UMIs maping on dissociation genes
totalSum <- Matrix::colSums(GetAssayData(object = wagner))
selection <- Data[genes, ]
selection[is.na(selection)] <- 0
dissociationSums <- colSums(selection)  
countSums <- merge(totalSum, dissociationSums, by="row.names", all=TRUE, sort= FALSE)
rownames(countSums) <- countSums$Row.names
countSums <- countSums[-1]
colnames(countSums) <- c("totalCount", "dissociationCounts")
countSums$percentage <- countSums$dissociationCounts/countSums$totalCount
## Save in meta.data of object
wagner@meta.data$percent.dissoc <- countSums$percentage

## Draw histogram for all samples
percentages <- wagner$percent.dissoc
hist(percentages, 
     breaks = 100, 
     col = "lightgrey", 
     main = paste("Expression dissociation-affected genes"), 
xlab = "Ratio of dissociation-affected genes to total gene count", ylab = "Number of cells", xlim = c(0, 0.20))
```

## Keep cells with dissociation percentages below the threshold of 6%
```{r}
wagner <- subset(x = wagner, 
                subset = percent.dissoc < 0.06)
dim(wagner)
table(wagner$orig.ident)
```

## Plot dataset
```{r}
DimPlot(object, 
        reduction="umap",
        group.by="RNA_snn_res.0.5")
DimPlot(object, 
        reduction="umap",
        group.by='old.ident')
```

## Batch correct with MNN
* Correct for c-sec vs grp effect
```{r}
object <- DietSeurat(object)
object <- RunFastMNN(object.list = SplitObject(object, split.by = "old.ident"))
```

## Redo clustering, UMAP calc with mnn correction
```{r}
object <- FindNeighbors(object, reduction = "mnn", dims = c(1:15))
object <- FindClusters(object, resolution = 0.3)
object <- RunUMAP(object, reduction = "mnn", dims = c(1:15))
DimPlot(object, 
        reduction = "umap")
DimPlot(object, 
       reduction = "umap",
       group.by = "old.ident")
saveRDS(object = object, file = "wagner_mnn.rds")

object[['seurat_clusters']] <- as.character(object@meta.data$seurat_clusters)
SaveH5Seurat(object, filename = "wagner_mnn.h5Seurat", overwrite=TRUE)
Convert("wagner_mnn.h5Seurat", dest = "h5ad", overwrite=TRUE)
```

## Remove Oocytes
```{r}
subset(x = object, subset = ZP3 < 3)
object <- subset(x = object, subset = ZP3 < 3)
object <- FindVariableFeatures(object, selection.method = "vst", nfeatures = 2500)
all.genes <- rownames(object)
object <- ScaleData(object, features = all.genes)
object <- RunPCA(object, features = VariableFeatures(object), nfeatures.print = 10)
object <- DietSeurat(object)
object <- RunFastMNN(object.list = SplitObject(object, split.by = "old.ident"))
object <- FindNeighbors(object, reduction = "mnn", dims = c(1:15))
object <- FindClusters(object, resolution = 0.3)
object <- RunUMAP(object, reduction = "mnn", dims = c(1:15))
DimPlot(object, 
        reduction = "umap",
        pt.size=1)
ggsave("umap_clusters.pdf")
DimPlot(object, 
       reduction = "umap",
       group.by = "old.ident",
       pt.size=1)
ggsave("umap_old_ident.pdf")

saveRDS(object = object, file = "wagner_mnn_no_oocytes.rds")

object[['seurat_clusters']] <- as.character(object@meta.data$seurat_clusters)
SaveH5Seurat(object, filename = "wagner_mnn_no_oocytes.h5Seurat", overwrite=TRUE)
Convert("wagner_mnn_no_oocytes.h5Seurat", dest = "h5ad", overwrite=TRUE)
```

## Calculate top markers
```{r}
markers <- FindAllMarkers(object,
                          only.pos = TRUE,
                          min.pct = 0.25,
                          logfc.threshold = 0.25)
markersPath = "markers_mnn_no_oocytes.tsv"
write.table(x = markers,
            file = markersPath,
            sep = "\t",
            quote = FALSE,
            row.names = FALSE)
markers %>%
    group_by(cluster) %>%
    slice_max(n = 4, order_by = avg_log2FC)

filtered_markers <- markers %>% filter(pct.1 > 0.6 & p_val_adj < 0.05)
write.table(x = filtered_markers,
            file = paste0(outputDir, "/filtered_markers_mnn_no_oocytes.tsv"),
            row.names = FALSE)

topMarkers <- filtered_markers %>%
              group_by(cluster) %>%
              top_n(5, avg_log2FC)
topMarkersPath = "topMarkers_mnn_filtered_markers.tsv"
write.table(x = topMarkers,
            file = topMarkersPath,
            sep = "\t",
            quote = FALSE,
            row.names = FALSE)
```

## Plot heatmap for top genes
* Notice there are no genes passing thresholds for cl 0. Confirm with cellxgene 
```{r}
## Calculate the per cluster mean expression
group_identity <- Idents(object=object)

normCounts <- GetAssayData(object=object)#, slot='scale.data')

## Substiture cell ID with cluster ID
colnames(normCounts) <- group_identity

## Remove cluster 0
# group_identity <- group_identity[group_identity!=0]
# group_identity <- droplevels(group_identity)

meanDF <- do.call(cbind, lapply(levels(group_identity), function(id){
   groupCounts <- normCounts[, colnames(normCounts) == id]
   df <- data.frame( c = apply(groupCounts, 1, mean))
   colnames(df) <- id
   return(df)
}))

meanDF_select <- as.matrix(meanDF[topMarkers$gene, ])
pdf("heatmap_top_markers.pdf")
heatmap.2(meanDF_select, 
          col="bluered", 
          trace = "none")
dev.off()
heatmap.2(meanDF_select, 
          col="bluered", 
          trace = "none")
```

## Top genes, separate clusters on old.ident
```{r}
old_ident_clusters <- paste(object[["old.ident"]]$"old.ident", object[["seurat_clusters"]]$seurat_clusters, sep="_" )
colnames(normCounts) <- old_ident_clusters
# group_identity <- as.factor(old_ident_clusters[old_ident_clusters != "grp_0" &  old_ident_clusters != "c_sec_0"])
group_identity <- as.factor(old_ident_clusters)

meanDF <- do.call(cbind, lapply(levels(group_identity), function(id){
   groupCounts <- normCounts[, colnames(normCounts) == id]
   df <- data.frame( c = apply(groupCounts, 1, mean))
   colnames(df) <- id
   return(df)
}))

meanDF_select <- as.matrix(meanDF[topMarkers$gene, ]) 
pdf("heatmap_top_markers_separ.pdf")
heatmap.2(meanDF_select, 
          col="bluered", 
          trace = "none")
dev.off()
heatmap.2(meanDF_select, 
          col="bluered", 
          trace = "none")
```

## Selected genes, separate clusters on old.ident
```{r}
selected_genes <- c("DCN", "FN1", "NR2F2", "ATF3",
                    "LUM", "COL1A1", "COL1A2", "COL3A1",
                    "COL6A1", "BGN", "HSPA1A")

meanDF_select <- as.matrix(meanDF[selected_genes, ])
pdf("heatmap_selected_markers_separ.pdf")
heatmap.2(meanDF_select, 
          col="bluered", 
          trace = "none")
dev.off()
heatmap.2(meanDF_select, 
          col="bluered", 
          trace = "none")
```

## Contrast s_sec_2 and grp_2, plot Volcano
```{r}
object[["old_ident_clusters"]] <- paste(object[["old.ident"]]$"old.ident", object[["seurat_clusters"]]$seurat_clusters, sep="_" )
deTab <- FindMarkers(object = object,
                     ident.1 = "c_sec_2", 
                     ident.2 = "grp_2",
                     group.by = 'old_ident_clusters',
                     min.pct = 0.25)

# Add a column to to the table store whether a gene is Up-regulated, Down-regulated or the change is Non-significant
# The levels of average log fold change and he adjusted P-value are used to decide on the above 
deTab$diffexpressed <- "Non-significant" 
deTab$diffexpressed[deTab$avg_log2FC > 0.5 & deTab$p_val_adj < 0.05] <- "Up"
deTab$diffexpressed[deTab$avg_log2FC < -0.5 & deTab$p_val_adj < 0.05] <- "Down"

 
deTab$delabel <- NA
deTab$delabel[deTab$p_val_adj<0.05] <- rownames(deTab)[deTab$p_val_adj<0.05]

markersPath = "markers_c_sec_2_vs_grp_2.tsv"
write.table(x = deTab,
            file = markersPath,
            sep = "\t",
            quote = FALSE,
            row.names = TRUE)

ggplot(data=deTab, aes(x=avg_log2FC, y=-log10(p_val_adj), col=diffexpressed, label=delabel)) +
        ggtitle("Volcano Plot") +
        geom_point() +
        theme_minimal() +
        geom_text_repel(max.overlaps = 15) +
        scale_color_manual(values=c("blue", "gray", "red")) +
        geom_vline(xintercept=c(-0.5, 0.5), col="red") +
        geom_hline(yintercept=-log10(0.05), col="red")
ggsave("c_sec_2_vs_grp_2.pdf")
```

## sessionInfo
```{r}
sessionInfo()
```
